---
title: "Novel Factor Target Gene Prediction through Multi-Omics Datasets"
subtitle: "OLET5602 Final Project"
author: "A. Kapoor, D. Langreiter, S. Udit, L. Richard"
date: "University of Sydney | `r format(Sys.time(), '%B, %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "assets/sydney-fonts.css", "assets/sydney.css", "assets/custom.css"]
    self_contained: true # if true, fonts will be stored locally
    nature:
      beforeInit: ["assets/remark-zoom.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # enable slide transitions by scrolling
---

## Introduction
--- 

.pull-left[

### Goals
  
.font90[.content-box-purple[
 **Main Goal:** To predict novel transcription factor target genes for Sox2 and Nanog using multi-omics datasets by applying classification techniques.
 
 **Secondary Goal:** Compare with Previous Studies 
]]

```{r load_data, message=FALSE, results='hide'}
 
 # Load necessary packages and data
load("Final_Project_ESC.RData", verbose = TRUE)

```

```{r setup, message=FALSE, results='hide', echo=FALSE}

suppressPackageStartupMessages({
  library(reshape2)
    library(e1071)
    library(ggplot2)
    library(ROCR)
    library(calibrate)
    library(dplyr)
    library(tibble)
    library(reshape2)
    library(kernlab)
    library(caret)
    library(pheatmap)
    library(randomForest)
    library(adabag)
    library(gbm)
    library(ggplotify)
    library(xgboost)
    library(nnet)
    library(RefManageR)
    library(pROC)
    library(doParallel)
    library(calibrate)
    library(vip)
})

set.seed(123)
```

]

.pull-right[

### Background

1. Transcriptional regulation is key to controlling gene expression during early embryonic development, guiding cell fate and differentiation.

2. Sox2 and Nanog are essential transcription factors that maintain pluripotency in embryonic stem cells. Advances in multi-omics technologies now enable us to explore these regulatory networks and discover new target gene
]



---

## Study Procedure (formerly Target Prediction and Validation)
--- 

- Only a subset of substrates of Sox/Nanog is known
- Develop a classification algorithm to predict additional substrates of Sox/Nanog
- Consider how to deal with potential imbalance class distribution. Evaluate and benchmark our predictions by comparing our predictions to the prediction in our previous study (Kim et al., 2020).


---
class: segue-yellow

## Data Description and Exporation

---

## Data Exploration
--- 

- Time course expression profiles 
- After induction of in vitro differentiation
- 3 Types of Omics data
  1. Transcriptome
  2. Proteome
  3. Epigenome - 6 Histone Markers
  
- Provided with a list of known Sox2/Nanog target Genes


---

## Transcriptome
--- 

```{r transcriptome_pca, message=FALSE, echo=FALSE}
# PCA analysis on the correlation matrix of the transcriptome data
cor.mat <- cor(Transcriptome)
pca.mat <- prcomp(cor.mat)

# Plot the PCA
grp <- rownames(pca.mat$x)
grp.col <- rainbow(nrow(pca.mat$x))
names(grp.col) <- rownames(pca.mat$x)

# Generate PCA plot
plot(pca.mat$x[,1], pca.mat$x[,2], col=grp.col[grp], pch=19, cex=2,
     xlab=paste0("PC1 (", round(summary(pca.mat)$importance[2,1]*100,1), "% variance)"),
     ylab=paste0("PC2 (", round(summary(pca.mat)$importance[2,2]*100,1), "% variance)"))

# Add sample labels to the plot
calibrate::textxy(pca.mat$x[,1], pca.mat$x[,2], labs=grp, cex=0.5)

```

---

## Proteome
--- 
```{r proteome_pca, message=FALSE, echo=FALSE}
cor.proteome <- cor(Proteome)
pca.proteome <- prcomp(cor.proteome)

# Using the previous correlation matrix and PCA results
cor.proteome <- cor(Proteome)
pca.proteome <- prcomp(cor.proteome)

# Get group labels and colors
grp <- rownames(pca.proteome$x)  # Set groups according to your data
grp.col <- rainbow(nrow(pca.proteome$x))
names(grp.col) <- rownames(pca.proteome$x)

# Plot the PCA
plot(pca.proteome$x[,1], pca.proteome$x[,2], col=grp.col[grp], pch=19, cex=2,
     xlab=paste0("PC1 (", round(summary(pca.proteome)$importance[2,1]*100,1), "% variance)"),
     ylab=paste0("PC2 (", round(summary(pca.proteome)$importance[2,2]*100,1), "% variance)"))

```

---

## Epigenome

### 6 Histone Markers
--- 

.imageFull[

```{r epigenome_pca, message=FALSE, echo=FALSE}

# Set up a 2x3 plotting area
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))  # Adjust margins as needed

# Function to perform PCA and plot
plot_pca <- function(data, title) {
  cor_data <- cor(data)
  pca <- prcomp(cor_data)
  
  grp <- rownames(pca$x)
  grp.col <- rainbow(length(grp))
  names(grp.col) <- grp
  
  plot(
    pca$x[, 1], pca$x[, 2],
    col = grp.col[grp],
    pch = 19, cex = 1.5,
    xlab = paste0("PC1 (", round(summary(pca)$importance[2, 1] * 100, 1), "% variance)"),
    ylab = paste0("PC2 (", round(summary(pca)$importance[2, 2] * 100, 1), "% variance)"),
    main = title
  )
  
  calibrate::textxy(pca$x[, 1], pca$x[, 2], labs = grp, cex = 0.5)
}

# Plot each dataset
plot_pca(H3K4me3, "H3K4me3")
plot_pca(H3K27me3, "H3K27me3")
plot_pca(H3K27ac, "H3K27ac")
plot_pca(H3K4me1, "H3K4me1")
plot_pca(H3K9me2, "H3K9me2")
plot_pca(PolII, "PolII")

# Reset to default single plot layout
par(mfrow = c(1, 1))
```
]

---
class: segue-yellow

## Target Gene Visualisation


---

## Heatmap I
--- 

.pull-left[
```{r heatmaps_one, message=FALSE, echo=FALSE}


# Top 100 most variable genes in the Transcriptome
top_genes_transcriptome <- Transcriptome[order(apply(Transcriptome, 1, var), decreasing = TRUE)[1:100], ]

# Heatmap for Transcriptome
pheatmap(top_genes_transcriptome, cluster_rows = TRUE, cluster_cols = TRUE, 
         show_rownames = FALSE, show_colnames = TRUE, 
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         main = "Heatmap of Top 100 Most Variable Genes in Transcriptome")

```
]

.pull-right[
- 1 - 6 hr genes exhibit lower levels of expression
- Suggest not many of top variable genes are not highly active immediately after differentiation.

]


---

## Heatmap II
--- 

.pull-left[
```{r heatmaps_two, message=FALSE, echo=FALSE}

# Select top 100 most variable genes from the Proteome
top_genes_proteome <- Proteome[order(apply(Proteome, 1, var), decreasing = TRUE)[1:100], ]

# Heatmap for Proteome
pheatmap(top_genes_proteome, cluster_rows = TRUE, cluster_cols = TRUE, 
         show_rownames = FALSE, show_colnames = TRUE, 
         color = colorRampPalette(c("blue", "white", "darkred"))(50),
         main = "Heatmap of Top 100 Most Variable Genes in Proteome")
```
]

.pull-right[
- 1 - 6 hr genes exhibit lower levels of expression
- Suggest not many of top variable genes are not highly active immediately after differentiation.

]

---

## Plot Over Time
### Time Course Expression
--- 
.pull-left[
```{r time_expr, message=FALSE, echo=FALSE, warning=FALSE}

# Subset key genes for visualization
key_genes <- c("SOX2", "NANOG") 

# Subset and melt Transcriptome data
key_genes_transcriptome <- Transcriptome[key_genes, ]
key_genes_transcriptome_melt <- melt(key_genes_transcriptome)
colnames(key_genes_transcriptome_melt) <- c("Gene", "Timepoint", "Expression")
key_genes_transcriptome_melt$DataType <- "Transcriptome"

# Subset and melt Proteome data
key_genes_proteome <- Proteome[key_genes, ]
key_genes_proteome_melt <- melt(key_genes_proteome)
colnames(key_genes_proteome_melt) <- c("Gene", "Timepoint", "Expression")
key_genes_proteome_melt$DataType <- "Proteome"

# Combine Transcriptome and Proteome data
combined_key_genes_melt <- rbind(key_genes_transcriptome_melt, key_genes_proteome_melt)

# Plot combined data
ggplot(combined_key_genes_melt, aes(x = Timepoint, y = Expression, color = Gene, group = Gene)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  facet_wrap(~DataType, scales = "free_y") +  
  theme_minimal() +
  labs(title = "Time-Course Expression of Key Genes in Transcriptome and Proteome",
       x = "Timepoint", y = "Expression Level") +
  scale_color_manual(values = c("SOX2" = "blue", "NANOG" = "red"))

```
]

.pull-right[
- Expression of SOX2 in the transcriptome starts slightly elevated.
- The Proteome shows SOX2 starting slightly positive however both then follow similar patterns of descent.
]

---
class: segue-yellow

## Classification Models


---

## SVM
--- 

.pull-left[

### Model Details

- Kernel (Radial Basis Function)
- Support Vectors (8064)
- Cost Parameter: C = 1 
- Gamma: 0.0159
- Training Accuracy: 87.79

]

.pull-right[
### Model Evaluation    

]

---

## Random Forest
--- 

.pull-left[

### Model Details

- Number of Trees: 1000
- Mtry: 5
- OOB (out-of-bag) Error rate: 24.42%
- Training Accuracy: 100%

]

.pull-right[
### Model Evaluation

]

---

## XGBoost
--- 

.pull-left[

### Model Details

- There wasn't much fluctuation
- Hypertuning the model would not improve performance
- High Cost at small marginal performance increase

]

.pull-right[
### Model Evaluation

]

---

## Neural Networks
--- 

.pull-left[

### Model Details

- TODO: Fill

]

.pull-right[
### Model Evaluation

]

---

## Bagged Cart
--- 

---
class: segue-yellow

## Discussion (TOPIC)


---

## Results
--- 

- **Final Choice of Model: XGB**
  - Slow Learner/Class Imbalance
- Predicts Approximately 2000
  - This is much higher than 400 from the previous study.
  - May be due to fewer data points or current downsampling.


---

## Future Work
--- 

- Class Imbalance Solutions
  1. Ensemble Methods
  2. Futher EDA to decide other means of removing data, or possibly leaving data imbalanced.

- Further Evaluation (Precision, Recall, F1-Score and AUC-ROC)


---
class: segue


## Thank You! (TOPIC)
